---
title: Biostats 597F
subtitle: Week 6 - Input and Output
output: ioslides_presentation
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(results="hide", eval=FALSE)
```

## Text file - readLines

- **readLines** read text file into a list, each component is text string of a line in the text
- It is usually used for unstructured text files when read.table can not be used

Example - Twitter text file

```{r}
umass <- readLines("umass_tt.txt", encoding = "UTF-8")
length(umass)
umass[[1]]
# percent of tweets are retweets
rt <- sapply(umass, function(x) grepl("^RT ", x))
mean(rt)
```

## Text file - readLines
Example - Remove some lines then read, e.g. SAS output

```{r}
f1 <- readLines("f1.txt")
# find and remove lines with ----
sep_lines <- sapply(f1, function(x) grepl("^----", x))
f1 <- f1[!sep_lines]
data <- read.table(text = paste(f1, collapse = "\n"), header = TRUE)
```

## Text file - read.table and read.csv

- Both read delimited text files to a data frame
- read.csv is a same as read.table with different default options such as comma separated file (csv) and with header row etc

Commonly used read options

- **header**: whether or not the file contains header row (variable names)
- **sep**: the delimiter usually "\\t", ",", " ", "|" etc
- **na.strings**: character value to be treated as missing
- **colClasses**: column classes
- **stringsAsFactors**: discussed before, should make it FALSE
- **text**: if read from a string instead of file

## Text file - read.table and read.csv

- read the data with different delimiter

```{r}
data_text <- "
Name|Age
Tom|45
John|23
"
data <- read.table(sep = "|", header = TRUE, stringsAsFactors = FALSE,
                   text = data_text)
str(data)
```

## Text file - read.table and read.csv

- specify different missing strings. In SAS, missing is represented by .

```{r}
data_text <- "
Name|Age
Tom|45
John|.
"
data <- read.table(sep = "|", header = TRUE, stringsAsFactors = FALSE,
                   text = data_text, na.strings = ".")
str(data)
```

## Text file - read.table and read.csv

- Specify variable type to read. By defualt, R automatically determine the variable type. 
- Choose from logical, integer, numeric, complex, character, raw, factor, Date, POSIXct
- Can input as a vector (will be recycled) or a named vector 
```{r}
data_text <- "
Name|BTHDATE|ZIP
Tom|1980-02-03|01003
John|1998-02-02|02446
"
data <- read.table(sep = "|", header = TRUE, stringsAsFactors = FALSE,
                   text = data_text,
                   colClasses = c("BTHDATE" = "Date",
                                  "ZIP" = "character"))
str(data)
```

## Text file - read.table and read.csv

- Even if a field contains a single non-numeric and non-empty character, the whole variable will be read as character
- After reading, we usually need to check if all variables are read and parsed correctly

```{r}
mtcars_large <- lapply(1:1000, function(x) mtcars)
mtcars_large <- do.call("rbind", mtcars_large)
write.csv(mtcars_large, file = "mtcars.csv", row.names = FALSE)
# modify any single value in the file and read
data <- read.csv("mtcars.csv", stringsAsFactors = FALSE)
```

## Large text file (less than momory)

- When file is large, but still loadable into memory, usually 100M - a few G, use **data.table** or **readr** for fast read

```{r}
CO2 <- CO2[rep(1:nrow(CO2), 30000), ]
write.csv(CO2, file = "CO2.csv", row.names = FALSE)
system.time(data <- read.csv("CO2.csv"))

library(data.table)
system.time(data <- fread("CO2.csv"))

library(readr)
system.time(data <- read_csv("CO2.csv"))
```

For more details of how to use those two functions in different sitations, google them.

## data.table and readr

- Both are much faster than read.csv, and data.table is usually faster than readr (usually less than 2 times faster)
- readr is developed by Hadley so it may better work with dply package, also developed by Hadley. We usually care less about the small loading speed gain by data.table
- fread automatically determine delimiter, but this is not useful in most cases
- Personally, I use readr and I have better experience using readr
- They determine variable type by looking at first n records

## Large text file not fitted to memory

We can use Apache Spark to process very large files. Spark has wrappter for R. 

http://spark.apache.org/ for apache spark including sparkR

http://spark.rstudio.com/ developed by RStudio to work with Spark in Rstudio

## Text file - JSON format

- JSON file format is very popular for web data
- Most results from Web API call are in JSON format
- **jsonlite** package is very handy to load JSON format data
- JSON is very similar to list

```{r}
library(httr)
library(jsonlite)
url <- "https://itunes.apple.com/search?term=taylor+swift&entity=musicVideo"
res <- content(GET(url))
# parse json data to R list
res <- fromJSON(res)
View(res[[2]])
```

jsonline automatically parse to data frame if the list can be converted to data frame

## Read text file from web

R can read data in web like a file

```{r}
data_link <- "http://archive.ics.uci.edu/ml/machine-learning-databases
/statlog/german/german.data"
data_link <- gsub("[[:space:]\n]", "", data_link)
data <- read.csv(data_link, header = FALSE, sep = " ",
                 stringsAsFactors = F)
```

## Output - writeLines, write.table, write.csv

- Reverse operation of readLines, read.table, read.csv

```{r}
writeLines(c("Line1", "Line 2", "Last line"), "test.txt")
write.csv(mtcars, "mt.csv", row.names = FALSE)
```

- Read document for different options such as na, quote etc

## Output - .rda file

- We can save any R object into a .rda and the load into R
- It works in the way to recover the work space
- It is usually to save model object together with functions and data to do pre-processing

```{r}
breaks <- c(0, 15, 25, Inf)
transform <- function(data) base::within(data, {mpg_c = cut(mpg, breaks)})
fit <- lm(cyl~mpg_c, data = transform(mtcars[-(1:5), ]))
save(fit, breaks, transform, file = "model.rda")
```

To predict on new data
```{r}
load("model.rda")
predict(fit, transform(mtcars[1:5, ]))
```

## Output - RDS file

R object can also be save as RDS file, and can be restored to a different name

```{r}
fit <- lm(cyl~mpg, data = mtcars)
saveRDS(fit, file = "model.rds")
```

```{r}
## readRDS to restore the object and assign to a name
fit_object <- readRDS("model.rds")
```

## rda/RData vs rds

- rda can contain multiple R objects and when loaded they are restored as what was saved (same object names). rds can only contain one R object, and can be restored and assigned to a different object name.
- We can use list to save multiple objects and then save as rds

