---
title: Biostats 597F
subtitle: Week 11 - Introduction to Rcpp
output: ioslides_presentation
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(results="hide", eval=FALSE)
```

## Rcpp

- Rcpp provides clean and approachable API to write high performance C++ code used by R
- Makes it easy to connect C++ and R, which used to be very painful
- You can use Rcpp with only little knowledge of C/C++

It is useful when

- Loops can not be easily vectorized
- Functions that will be called millions of times, a slight improvement by rewriting in C++ can be very beneficial
- Problems that need complex data structures

## Installation

- Windows: Install Rtools
- Mac OSX: Install XCode for gcc

- Install Rcpp package

## Simple Example

We can create a sample Cpp file for Rcpp using Rstudio

- File >> New File >> C++ file
- Save the sample file
- Examine the basic component of Cpp file
- Use the function

```{r}
library(Rcpp)
sourceCpp("example1.cpp")

timesTwo(123)
```

## Simple Example

- Need to specify data type for each variable and output
- There is a corresponding variable type in Rcpp for each R variable type, e.g. NumericVector is corresponding to a numeric vector. NumeriMatrix for numeric matrix
- Need to put ```// [[Rcpp::export]]``` before the function inorder to use it in R

## Example using Vector

- When use for loop, index starts from 0 in C
- We can use x.size() to know the length of the vector
- Must use `=` for assignment

Function to calculate mean of a vector

```{r}
library(microbenchmark)
sourceCpp("examples.cpp")

microbenchmark(mean(rnorm(100000)),
               meanC(rnorm(100000)))
```
## Example Using Vector

Cumulative sum of positive values in the vector

```{r}
sourceCpp("examples.cpp")
# R version
csum <- function(x) {
  sum(x[x > 0])
} 

microbenchmark(
  csum(rnorm(100000)),
  csumC(rnorm(100000))
)

```
## Vector Example

Rcpp can operate or R object inplace, e.g. replace elements of x without recreating a new vector.

Replace negative values in x to be 0

```{r}
sourceCpp("examples.cpp")
# R version
fill0 <- function(x) {
  x[x < 0] <- 0
  x
} 

x <- fill0(c(1, 2, -3))
fill0(x)
fill0C(x) ## nothing return, but x is changed

```
## Matrix

- Mostly used are NumericMatrix and IntegerMatrix
- Use X(i, j) to extract matrix elements
- Use X.nrow() and X.ncol() to get the dimension of the matrix

Example: row sum of a matrix

```{r}
sourceCpp("examples.cpp")

m <- matrix(sample(1:10, 100, replace = TRUE), nrow = 10)

rowSums(m)  # R version
rowSumC(m)
```
## Other Data Types

- Many other data types are available in Rcpp such as list, data frame, function etc
- However vector and matrix are most often used 
- For more details we can refer to 

http://adv-r.had.co.nz/Rcpp.html and the book **Seamless R and C++ Integration with Rcpp**


## Example - Model Prediction

A score to predict someone to receive vaccine or not is computed based on three variables: age, gender (female or not), ily (whether was infected by virus in last year or not) 

The probability is computed by

```{r}
vacc1a <- function(age, female, ily) {
  p <- 0.25 + 0.3 * 1 / (1 - exp(0.04 * age)) + 0.1 * ily
  p <- p * if (female) 1.25 else 0.75
  p <- max(0, p)
  p <- min(1, p)
  p
}
```

However this does not take vector as input

## Example - Model Prediction

Vectorized input version

```{r}
vacc1 <- function(age, female, ily) {
  n <- length(age)
  out <- numeric(n)
  for (i in seq_len(n)) {
    out[i] <- vacc1a(age[i], female[i], ily[i])
  }
  out
}
```

## Example - Model Prediction

Improved R vectorized version

```{r}
vacc2 <- function(age, female, ily) {
  p <- 0.25 + 0.3 * 1 / (1 - exp(0.04 * age)) + 0.1 * ily
  p <- p * ifelse(female, 1.25, 0.75)
  p <- pmax(0, p)
  p <- pmin(1, p)
  p
}
```

## Example - Model Prediction

Rcpp version (see examples.R)

Comparison

```{r}
sourceCpp("examples.cpp")
n <- 1000
age <- rnorm(n, mean = 50, sd = 10)
female <- sample(c(T, F), n, rep = TRUE)
ily <- sample(c(T, F), n, prob = c(0.8, 0.2), rep = TRUE)

stopifnot(
  all.equal(vacc1(age, female, ily), vacc2(age, female, ily)),
  all.equal(vacc1(age, female, ily), vacc3(age, female, ily))
)

microbenchmark(
  vacc1 = vacc1(age, female, ily),
  vacc2 = vacc2(age, female, ily),
  vacc3 = vacc3(age, female, ily)
)

```





