---
title: Biostats 597F
subtitle: Week 6
output: ioslides_presentation
---

## Object Oriented Programming

What happens when we type a name in the console?

```{r, results='hide'}
x <- 1:10
x

fit <- lm(mpg~gear + cyl, data = mtcars)
fit

```

- It will run **print** function on the object
- According the class of the object, the print function will be different. (try to modify the class of the fit to be **list** to see what happens)
- Example in predictive modeling, **predict** function works for different models such as random forest, linear model, logistic model.

## Object Oriented Programming

Everything in R is object. 

- R promotes encapsulation
- R classes are polymorphic. Same call to different objects may trigger different actions depending on the class of the object
- R allows inheritance to extend class. One example is generalized linear model object inherits linear model object and extends on that
- Two types of classes are available in R S3 and S4
- S3 is simple and flexible but not rigrous, most R objects are S3 class. S4 is more rigrous 
```{r, results='hide'}
isS4(fit)
isS4(1:5)
```

## Create S3 Class

- No need to define an S3 class, just assign the class to the class attributes to the object (usually a list)
- Although there are expectations on the data structure for a class, R does not do any check.

We can look at ```lm``` function to see how assign class to an object
```
print(lm)
```

Example of our own class

```
x <- list(name = "John", salary = 5000)
class(x) <- "employee"
```

## Create S3 Class

We can add/remove/modify componets from an S3 class, R will not complain until we call a function using the affected element

```{r, results="hide", eval=FALSE}
fit <- lm(mpg ~ cyl + gear + wt, data = mtcars)
names(fit) # what componnents in the object
fit$who <- "Xiangdong" # add one element
fit$call <- "nothing"  # type fit to see the change
fit$fitted.values <- NULL # remove one element
summary(fit) # still work
print(fit) # still work
fit$rank <- NULL # are summary, print still working?
```

When we train a model and save the output, we can append addtional components to the model object such as version number. You can not do this for S4 class.

## Polymorphic - Generic function

Polymorphic - When a generic function, such as print, summary, is called on an object, the actual function call is different for different classes of object.

Here is how it works as an example for print(fit), fit has class **fit**

- Search for function **print.lm**
- If it is available then call **print.lm(fit)**
- Otherwise call **print.default(fit)**

To see the definition of print.lm
```
stats:::print.lm
```

We add stats::: because the function is in stats package and not exported

## Generic function

- Generic function is the function that will dispatch different functions to different classes
- List all generic functions for a class
```
methods(class = "lm")
```
- List all available methods for a generic function
```
methods("print")
```
- How generic function is defined
```
print(print)
```
There is only one line code. **UseMethod** is the dispatch function to find the appropriate function to call.

## Example

We can create our own student class

generate data
```{r, results='hide', eval=FALSE}
me <- list(first = "xiangdong",
           last = "gu",
           age = 30,
           department = "biostat",
           degree = "phd")
```

specify class
```{r, results='hide', eval=FALSE}
class(me) <- "student"
```

## Example
define method

```{r, results='hide', eval=FALSE}
print.student <- function(x) {
  s <- sprintf("%s %s, %s years old, is a %s in %s department",
               x$first, x$last, x$age, x$degree, x$department)
  cat(s)
}
```

```{r, results='hide', eval=FALSE}
summary.student <- function(x) {
  s <- ifelse(x$age > 28, "old", "young")
  cat(sprintf("%s %s is %s", x$first, x$last, s))
}
```

We can implement as many as possible methods for this class as long as the generic function is already defined.

## User-defined generic function

We can define our own generic function, similar to **print**

```{r, results='hide', eval=FALSE}
getage <- function(x, ...) UseMethod("getage", x)
getage.student <- function(x) x$age
getage(me)
```

This is more useful if we have multiple classes we want to have **getage** method.

```{r, results='hide', eval=FALSE}
j <- list(name = "Joe", empage = 45)
class(j) <- "employee"
getage.employee <- function(x) x$empage
getage(j)
```

With class and methods, we get a unified interface of calling functions.

## Inheritance

If an object class A extends from another object class B, we can reuse what has been implemented in class B.

Example - A new class **phdstudent** for phd student inheriting from **student**

```{r, results='hide', eval=FALSE}
x <- list(first = "xiangdong",
          last = "gu",
          age = 30,
          department = "biostat",
          degree = "phd",
          advisor = "raji")
class(x) <- c("phdstudent", "student")
print.phdstudent <- function(x) {
  cat(sprintf("advisor is %s", x$advisor))
}
x
summary(x)
```

## Inheritance

Basically, it first searches for **print.phdstudent** funcition, if not found, then search for **print.student** function.

## Double dispatch

We can define methods for operators such as "+"

Example: use + to add number of rows of data frames
```{r, results="hide", eval=FALSE}
`+.data.frame` <- function(e1, e2) {
  nrow(e1) + nrow(e2)
}
mtcars + iris
```

This is used ggplot2 package, the class of ggplot is **gg**

```
library(ggplot2)
methods(class = "gg")
ggplot2:::`+.gg`
```

