---
title: Biostats 597F
subtitle: Week 9 - Efficient R Programming
output: ioslides_presentation
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(results="hide", eval=FALSE)
```

## Efficient R Prgramming

- R can be very slow if not used properly
- Should avoid R programming pitfalls leading to poor performance
- Vectorize as much as possible. Many problems can be vectorized enven it does not like so
- Understand memory usage by R helps to understand why code is slow
- Use Rcpp package to rewrite the step that takes most of the computation time if it can save time significantly

## Performance Measurement

**System.time** funcion can be used to measure how long a piece of code run.

```{r}
system.time(y <- rnorm(1000000))
```

When the running is too small, this function is less precise. Instead we can use **microbenchmark** to get a more precise measurement.

```{r}
install.packages("microbenchmark")
library(microbenchmark)

microbenchmark(
  rnorm(1000),
  rnorm(10000)
)
```

## Vector Assignment

- Most R objects are **immutable**. That is, if we modify a vector the whole vector will be changed and a new vector is created. This may have huge performance impact.

```{r}
x1 <- rnorm(1e4)
x2 <- rnorm(1e6)
x3 <- rnorm(1e7)
microbenchmark(
  x1[1] <- 0,
  x2[1] <- 0,
  x3[1] <- 0
)
```
- The reason it takes much longer for a long vector is R will create a copy of the vector, then make the change, and then assign x to the new vectork.

## Vector Assignment

Now we understand why we should not do assigning values one by one

```{r}
x <- rnorm(10000)
addone1 <- function(x) {
  for (i in 1:length(x)) x[i] <- x[i] + 1
  x
}
addone2 <- function(x) x + 1
microbenchmark(
  addone1(x),
  addone2(x)
)
```

**addone2**, the vectorized version, does not result numerous copy of the vector.

## Copy-on-change

```{r}
z <- runif(10)
y <- z
tracemem(z)
tracemem(y)
y[1] <- 0
tracemem(z)
tracemem(y)
```
- Initially y binds to same memory as z after assignment
- When y is changed, a new copy is created and y pointed to the new address. z has no change

## Pre-allocate memory

```{r}
method1 = function(n) {
  vec = NULL # Or vec = c()
  for(i in seq_len(n))
    vec = c(vec, i)
  vec
}
method2 = function(n) {
  vec = numeric(n)
  for(i in seq_len(n))
    vec[i] = i
  vec
}
method3 = function(n) seq_len(n)
microbenchmark(method1(n), method2(n), method3(n))
```

## Vectorized Programming

- We should avoid unnecessary assigning by using vectorized programming
- Minimize function calls will help code run faster

**Excercise**

Urn 1 contains ten blue marbles and eight yellow ones. In urn 2, the mixture is six blue and six yellow. We draw a marble at random from urn 1, transfer it to urn 2, and then draw a marble at random from urn 2. What is the probability that that second marble is blue? 

## ifelse may not be efficient

ifelse is useful but it may be slow

- example missing value imputation

```{r}
f1 <- function(x) ifelse(is.na(x), 0, x)
f2 <- function(x) {
  x[is.na(x)] <- 0
  x
}
x <- sample(c(1, 2, NA), 100000, replace = TRUE)
microbenchmark(f1(x),
               f2(x))
```

## pmin and pmax

pmin and pmax obtains the pairwise min and max between two vectors

- Example: Federal tax rate calculation for Single

see https://www.irs.com/articles/2016-federal-tax-rates-personal-exemptions-and-standard-deductions

```{r}
breaks <- c(0, 9275, 37650, 91150, 190151, 413350, 415050)
rate <- c(0.1, 0.15, 0.25, 0.28, 0.33, 0.35, 0.396)
getFedTax <- function(income) {
  diffs <- diff(c(breaks, Inf))
  exposure <- income - breaks
  exposure <- pmax(pmin(exposure, diffs), 0)
  sum(exposure * rate)
}
getFedTax(80000)
```

## outer function

**outer** function runs function on each value pairs between vector x and y

- Example: for each value in x, computer 1-10 powers as a matrix

```{r}
x <- 1:5
y <- 1:10
outer(x, y, "^")
```
- Example: generate vector of a1-a10, b1-b10,...z1-z10
```{r}
f <- function(x, y) paste0(x, y)
v <- outer(letters, 1:10, "f")
as.vector(v)
```
## expand.grid function

**expand.grid** can be used to enumerate all the possible combinations

- Example: generate all possible friendship pairs between a few names

```{r}
nms <- c("Tom", "Alice", "Kim", "Brenda")
expand.grid(nms, nms)
```

- Example: generate a longitudinal data for a group of ids
```{r}
id <- 1:10
date <- as.Date(sprintf("2016-%s-01", 1:12))
data <- expand.grid(date = date, id = id)
```

